---
title: "loanPredictionModel"
author: "Kanishk Kapoor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("dplyr")
library("tidyr")
library("caret")
library("rpart")
library("rpart.plot")
library("ggplot2")
library("car")
library("ROSE")
library("randomForest")
library("pROC")
```

### Data Loading & Cleaning
```{r}
client_data <- read.csv("bankloans.csv",stringsAsFactors = FALSE)
```

```{r}
head(client_data)
summary(client_data)
str(client_data)
```

```{r}
colSums(is.na(client_data))
```

### Data Processing

```{r}
client_data$default <- as.factor(client_data$default)
```
```{r}
client_data[, c("income", "debtinc", "creddebt", "othdebt")] <- scale(client_data[,c("income", "debtinc", "creddebt", "othdebt")])
```

```{r}
str(client_data)
```
### T-Test
```{r}
defaulters <- client_data %>% filter( default == 1)
non_defaulters <- client_data %>% filter( default == 0)
```

```{r}
# T-test for income
t_income <- t.test(defaulters$income, non_defaulters$income, var.equal = TRUE)
print(t_income)

#T test for debt to income ratio
t_debtinc <- t.test(defaulters$debtinc, non_defaulters$debtinc, var.equal = TRUE)
print(t_debtinc)

#T test for credit to debt ratio
t_creddebt <- t.test(defaulters$creddebt, non_defaulters$creddebt, var.equal = TRUE)
print(t_creddebt)

#T test for other debt
t_othdebt <- t.test(defaulters$othdebt, non_defaulters$othdebt, var.equal = TRUE)
print(t_othdebt)

```
1️⃣ Income
	•	t = -1.8797, p-value = 0.06056
	•	p-value > 0.05 → Not statistically significant
✅ Conclusion: Income may not be a strong predictor for loan default in our dataset.

2️⃣ Debt-to-Income Ratio (debtinc)
	•	t = 11.175, p-value < 2.2e-16
	•	p-value < 0.05 → Highly significant
✅ Conclusion: Debt-to-income ratio is a strong predictor of default.

3️⃣ Credit-to-Debt Ratio (creddebt)
	•	t = 6.6688, p-value = 5.249e-11
	•	p-value < 0.05 → Highly significant
✅ Conclusion: Credit-to-debt ratio is also a strong predictor of default.

4️⃣ Other Debts (othdebt)
	•	t = 3.8912, p-value = 0.0001093
	•	p-value < 0.05 → Significant
✅ Conclusion: Other debts also contribute to loan default risk.

After conducting a T-test, we identified significant predictors (debtinc, creddebt, and othdebt).
However, we need to check if these predictors are highly correlated (multicollinear), which can negatively impact model performance.

This is where Variance Inflation Factor (VIF) comes in.

⸻
 What is VIF?
	•	VIF measures multicollinearity between independent variables.
	•	A high VIF (> 5 or 10) means a variable is highly correlated with others, leading to redundancy in the model.
	•	Why is this bad?
	•	Unstable model coefficients
	•	Incorrect predictor importance
	•	Overfitting issues
	
```{r}
client_data$default <- as.numeric(client_data$default)
vif_model <- lm(default ~ age + ed + employ + address + income + debtinc + creddebt + othdebt, data= client_data)

vif_values <- vif(vif_model)
print(vif_values)
```

VIF < 5 → No multicollinearity

### Model Building

Before building a model, we need to split our dataset into two parts:

1️⃣ Training Set (80%) → Used to train the model.
2️⃣ Testing Set (20%) → Used to evaluate the model’s performance.

When building machine learning models, we need to ensure that:
	•	The model learns patterns from the training data.
	•	The model is tested on unseen data (test set) to check its real-world performance.
	•	This prevents overfitting, where the model memorizes the data instead of generalizing it.
	
	Even though we are splitting 80% for training and 20% for testing, the selection of which rows go into train vs. test is random.
	
	 Why is Random Splitting Important?
	•	If we just take the first 80% of rows, we might introduce bias (e.g., all low-income customers might be in training, and high-income ones in testing).
	•	Random splitting ensures that the training set and test set represent the full dataset fairly.
	
	The function set.seed(42) ensures that random processes (like splitting data) always give the same result when you run the code multiple times.
	•	Without set.seed(): Every time you run the code, a different random split occurs.
	•	With set.seed(42): The same random split is used, ensuring consistent results.


⸻
```{r}
set.seed(30) #ensuring reproducibility
trainIndex <- createDataPartition(client_data$default, p=0.8, list = FALSE)
trainData <- client_data[trainIndex, ]
testData <- client_data[-trainIndex, ]
```

#### Logistic Regression Model
```{r}
log_model <- glm(default ~ age + ed + employ + address + debtinc + creddebt + othdebt, data= trainData, family = binomial)

summary(log_model)
```
Variable Estimate
Intercept (-0.3325)
Baseline log-odds of default when all predictors are 0.
age (+0.0505)
Older customers have a slightly higher risk of default (p = 0.009)  Significant.
ed (+0.1020)
Education level does not significantly impact default (p = 0.467) Not significant.
employ (-0.2689)
More years of employment reduces the probability of default (p < 0.001) Strongly significant.
address (-0.1443)
More years at the same address reduces default risk (p < 0.001) Strongly significant.
debtinc (+0.6634)
Higher debt-to-income ratio increases default risk (p < 0.001)  Strong predictor.
creddebt (+1.2475)
Higher credit-to-debt ratio significantly increases default risk (p < 0.001)  Strong predictor.
othdebt (-0.1168)
Other debts do not significantly affect default (p = 0.635)  Not significant.

### Model Retrain by omitting insignificant variables

```{r}
log_model_updated <- glm(default ~ age + employ + address + debtinc + creddebt, data = trainData, family = binomial)

summary(log_model_updated)
```
```{r}
AIC(log_model)
AIC(log_model_updated)
```

New model has the low AIC score which implies that the new model is better.

### Predict on Test Data
```{r}
pred_prob <- predict(log_model_updated, testData, type = "response")

cut_off <- 0.5
pred_class <- ifelse(pred_prob > cut_off, 1 , 0)

pred_class <- as.factor(pred_class)
```


### Model Performance Evaluation using Confusion Matrix

```{r}
# Convert actual values to factor
testData$default <- as.factor(testData$default)

# Convert predicted values to factor with same levels
pred_class <- factor(pred_class, levels = levels(testData$default))
conf_matrix <- confusionMatrix(pred_class, testData$default, positive ="1")

print(conf_matrix)
```
```{r}
table(trainData$default)
table(testData$default)
```

✅ The dataset is imbalanced since the number of defaulters (1) is much higher than non-defaulters (2).
✅ The model is likely biased towards predicting default (1), which explains why our confusion matrix showed 100% Sensitivity but 0% Specificity (predicting only 1).
📌 How to Handle Class Imbalance?

1️⃣ Resample the Data (SMOTE or Undersampling/ Oversampling)
2️⃣ Change Cut-off Threshold (e.g., from 0.5 to 0.3 or 0.4)
3️⃣ Try a Different Model (Random Forest or Decision Tree)


SMOTE (Synthetic Minority Oversampling Technique) generates synthetic examples for the minority class (non-defaulters) to balance the dataset.

```{r}
# Convert default column to factor
trainData$default <- as.factor(trainData$default)

# Apply SMOTE (Over-samples minority class, under-samples majority)
set.seed(30)  # Ensures reproducibility
trainData_balanced <- ROSE(default ~ ., data = trainData, seed =30)$data

# Check new class distribution
table(trainData_balanced$default)

# Check new class distribution
table(trainData_balanced$default)
```
ROSE automatically balances the classes by generating synthetic samples.

### Retrain Logistic Regression on Balanced Data

```{r}
log_model_balanced <- glm(default ~ age + employ + address + debtinc + creddebt , data= trainData_balanced, family = binomial)

summary(log_model_balanced)
```
```{r}
table(trainData$default)
```

```{r}
# Predict probabilities
pred_prob_balanced <- predict(log_model_balanced, testData, type = "response")

# Convert probabilities to class labels using cutoff = 0.5
pred_class_balanced <- ifelse(pred_prob_balanced > 0.5, 1, 2)  # "2" represents non-defaulter

# Convert to factor
pred_class_balanced <- as.factor(pred_class_balanced)

# Ensure factor levels match test data
pred_class_balanced <- factor(pred_class_balanced, levels = levels(testData$default))
```

```{r}
# Compute confusion matrix
conf_matrix_balanced <- confusionMatrix(pred_class_balanced, testData$default, positive = "1")

# Print results
print(conf_matrix_balanced)
```

Poor Model Performance, will try decision tree model for further evaluation.

### Decision Tree Classifier Model
```{r}

tree_model <- rpart(default ~ age + employ + address + debtinc + creddebt, data = trainData, method = "class")

# Predict on test data
pred_tree <- predict(tree_model, testData, type = "class")

# Evaluate performance
conf_matrix_tree <- confusionMatrix(pred_tree, testData$default, positive = "1")
print(conf_matrix_tree)
```
Tree model significantly improved accuracy (78.42%) and recall (87.38%). However, specificity (52.78%) is still low, meaning the model struggles to correctly classify non-defaulters. Decision Trees can naturally handle imbalanced datasets without needing SMOTE or ROSE.

```{r}
colSums(is.na(trainData))


```


### Random Forest Model
What is Random Forest?

✅ Random Forest = Many Decision Trees Working Together
	•	Instead of a single decision tree, it creates multiple decision trees.
	•	Each tree learns from a random subset of data (called bootstrapping).
	•	At the end, it aggregates predictions from all trees (majority voting for classification, average for regression).

📌 Key Idea: “A group of weak models together form a strong model.”

✅ Best for Large, Complex Datasets
	•	Handles missing values well (unlike Logistic Regression).
	•	Works with both categorical & numerical data.
	•	Great for imbalanced datasets (it captures rare patterns well).
	•	Prevents overfitting (due to averaging multiple trees).

🔴 Not Ideal When:
	•	You need a highly interpretable model (Random Forest is complex).
	•	The dataset is very small (a simple Decision Tree may work just as well).
```{r}
set.seed(30)

rf_model_orignal <- randomForest(default ~ age + employ + address + debtinc + creddebt, data = trainData, ntree = 500, mtry = 3, importance = TRUE)

pred_rf_orignal <- predict(rf_model_orignal, testData)

```

```{r}
print(pred_rf_orignal)
```

```{r}

# Evaluate performance
conf_matrix_rf_original <- confusionMatrix(pred_rf_orignal, testData$default, positive = "1")

# Print results
print(conf_matrix_rf_original)
```
Our Random Forest model outperforms both Decision Tree and Logistic Regression, achieving:
✅ Higher accuracy (79.86%)
✅ Strong sensitivity (92.23%) → Great at catching defaulters
✅ Balanced accuracy of 68.34% → Better than before

### Visualization
```{r}
varImpPlot(rf_model_orignal)
```
```{r}
str(pred_rf_orignal)

```

```{r}
pred_rf_orignal <- as.numeric(pred_rf_orignal)

roc_curve <- roc(testData$default, pred_rf_orignal)
plot(roc_curve, col="blue", main="ROC Curve - Random Forest Model")
```
```{r}

ggplot(data.frame(probability=pred_rf_orignal, default=testData$default), aes(x=probability, fill=factor(default))) +
  geom_histogram(alpha=0.5, bins=20, position="identity") +
  labs(title="Predicted Default Probability Distribution", x="Probability of Default", fill="Actual Class") +
  theme_minimal()
```

